Peyton Donaldson
title: "Car Thefts"
9/23/26

Follow along with the demo to learn how to calculate standard deviation and how to create a histogram of the data using R and the Tidyverse. You should also open the CarThefts_Explainer.Rmd file and use that as a reference.

Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the car thefts dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv")
```

### Task 3: Calculate the standard deviation of the number of car thefts, along with the mean

```{r}
car_thefts |> summarize(mean = mean(count), sd = sd(count))
```


### Task 4: Create a histogram of the number of car thefts

```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2) +
  geom_vline(aes(xintercept = mean(count)), color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(aes(xintercept = mean(count) - sd(count)), color = "blue", linetype = "dashed", linewidth = 1) +
  geom_vline(aes(xintercept = mean(count) + sd(count)), color = "blue", linetype = "dashed", linewidth = 1)
```


### Task 5: Add a line to the histogram that shows the mean of the number of car thefts, and two other lines to show one standard deviation in both directions


### Task 6: Describe your results

Is the distribution normal or skewed in one direction? Write a sentence in which you describe the range of the data covered by one standard deviation from the mean on both sides - how much of the data falls in that range?

The distribution is skewed a bit to the right, but still somewhat normal. The range of data is averaged between 12.6 + 4.3 and 12.6 - 4.3, which the red line represents the mean and the blue represents one sd on either side. However, when reading this histogram, most but not all points fall between 8 & 17, which is a more accurate way to represent this model to readers. In regards to data, in that range, I'd represent this data by saying there's a medium in between 8 & 17 thefts for the month of January 2024.  


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
