Peyton Donaldson
title: "Polling Simulation: Sampling and the Central Limit Theorem"
10/7/26

In this activity, you'll use data from the 2024 Cooperative Election Study (CES) — a large-scale national survey conducted during the 2024 election — to simulate what happens when a news organization tries to estimate public opinion from a poll sample. The full dataset has nearly 55,000 respondents, which we'll treat as our population.

Your goal is to see firsthand how sample size affects the accuracy of estimates, and what the distribution of sample estimates actually looks like — the core idea behind the Central Limit Theorem.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from this URL and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

We'll focus on one question: what share of American adults identify as religiously unaffiliated (atheist, agnostic, or nothing in particular)? This has been a major story in American religion journalism for years.

First, we'll create a new column called `unaffiliated` that is `TRUE` if a respondent's `trad` value is `"Atheist"`, `"Agnostic"`, or `"Nothing in Particular"`, and `FALSE` otherwise. Then calculate the proportion of the full dataset that is unaffiliated. Save this value as `population_proportion`.

```{r}
ces <- ces |>
  mutate(unaffiliated = trad %in% c("Atheist", "Agnostic", "Nothing in Particular"))

proportion <- mean(ces$unaffiliated)
```

That population proportion is the "true" population value that any poll is trying to estimate. Write it down — you'll be comparing everything else to it.

### Task 4: Take a single small sample

Use `set.seed(42)` and then sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of unaffiliated respondents in your sample.

```{r}
set.seed(42)

sample_50 <- ces |>
  sample_n(50)

sample_50 |>
  summarize(proportion = sum(unaffiliated) / 50)
```

How close is your sample estimate to the original proportion? Is it surprising?

I'd say the sample estimate of 0.32 is pretty close to the original proportion of 0.35985, with a difference of about 4 percentage points. This isn’t too surprising because I expected a sample of only 50 people will have some variation from the full dataset.I think if I used a larger number it'd be closer!

### Task 5: Take a single larger sample

Now sample 500 rows. Set a different seed value, grab the sample and save it as `sample_500`. Calculate the proportion of unaffiliated respondents in this sample.

```{r}
set.seed(50)

sample_500 <- ces |>
  sample_n(500)

sample_500 |>
  summarize(proportion = sum(unaffiliated) / 500)
```

How did the estimate change compared to the sample of 50? What does this suggest about sample size?

The estimate increased from 0.32 with the sample of 50 to 0.352 with the sample of 500. The larger sample is much closer to the original proportion of 0.35985, so I'd say that larger samples tend to give more accurate estimates because they are less affected by the random variation.

### Task 6: Simulate many polls

A single sample can get lucky or unlucky. To understand how polling actually works, we need to see the *distribution* of estimates across many samples. The code below draws 1,000 samples of size 50, calculates the proportion of unaffiliated respondents in each, and stores the results.

```{r}
set.seed(50)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$unaffiliated)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side — one for the 1,000 samples of size 50 and one for the 1,000 samples of size 500. Add a vertical line at `proportion` to each so you can see how well the samples are centering on the true value.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share Religiously Unaffiliated",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**What shape do both histograms have?** Describe them in plain language. Does it surprise you that this shape emerges from a variable that is simply TRUE or FALSE for each respondent?

I'd say both histograms have a bell-shaped pattern, with most of the sample estimates clustered around the true population proportion and fewer estimates farther away. The sample of 500 is much narrower and more concentrated than the sample of 50. It is a little surprising that this shape comes from a variable that is just TRUE or FALSE, but when we take many random samples, I think the proportions start to form this bell-shaped distribution.

**Compare the two histograms.** What is the most visible difference between the n=50 and n=500 distributions? What does that difference tell you about what happens when a poll uses a larger sample?

The biggest difference is that the n=500 histogram is much narrower and more concentrated around the true proportion, while the n=50 histogram is more spread out. This represents that polls with larger samples give more consistent and accurate estimates, with less variation from different samples.

**Both distributions are centered near the red line.** What does it mean that 1,000 different random samples — each one drawn independently — tend to cluster around the same value? What would it mean for journalism if that *weren't* the case?

This shows that even though every random sample has different people, the results still tend to stay close to the true population proportion. That makes polls useful because journalists can use a smaller sample to get a good idea of what a much larger population looks like. If this wasn’t the case, poll results could be all over the place, making them much harder for journalists to trust and report on.

**A reporter tells you their story is based on a poll of 50 people.** Based on what you've seen in the simulation, what questions would you want to ask before trusting that number? You don't need to use any technical terms — just describe what you observed.

Before trusting the number, I would want to know how those 50 people were chosen and whether they actually represent the larger group the story is talking about. I would also want to know if a larger poll found similar results. From the simulation, we saw that samples of 50 can vary a lot more, so I wouldn’t want to make a big conclusion based on one small sample!

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
