Peyton Donaldson
title: "Sampling"
9/30/26

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library (tidyverse)
```

### Task 2: Load 1,000 rows of data with the ages of people from 18-80
```{r}
ages <- data.frame(
  age = sample(18:80, 1000, replace = TRUE))
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sample_ages <- ages |>
  sample_n(100)
```


### Task 4: Calculate the mean age and standard deviation of the sample
```{r}
sample_ages |>
  summarize(mean_age = mean(age),sd_age = sd(age))
```


### Task 5: Calculate the mean age and standard deviation of the population
```{r}
ages |>
  summarize(mean_age = mean(age),sd_age = sd(age))
```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The sample and population statistics are pretty similar I think, overall. But there are still some noticeable differences, like the population had a mean age of 48.49 and a standard deviation of 18.07, while the sample had a slightly higher mean age of 50.97 and a standard deviation of 18.82. This makes sense because the sample only includes 100 people from the full population of 1,000. So, if I used an even smaller sample size, I would expect the mean and standard deviation to vary more because fewer ages are being represented, making the sample less likely to closely match the full population as is. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
