---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```




### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```

### Task 3: Calculate the percentage change in each category between Fall 2026 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2026`- `Fall 2021`)/`Fall 2021` * 100)
```



### Task 4: Describe the results

Write a few sentences about what you found in those percentages. What's the story here? What questions do you have?

In the percentages, there's a massive/infinite "increase" for the student counseling center, which I find a bit concerning or confusing for the campus, and how it has increased so much over the years. Is it due to marketing, branding, or are UMD students going through considerably immeasurable issues?

Also, for the shuttle bus percentage change of 58% and knowing there is a new change in electric buses how have the past few years of traditional gas-powered buses affected the environment, campus, or students to want to increase fees?

Lastly, why is athletics so consistent over the years? Why does UMD campus feel the need to keep funding it?  


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
