Peyton Donaldson
title: "HW2 Prince George's County Crime"
9/9/26

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

Based off a quick glance, the constant "average" or consistency for the homicide and sex offense category is a little concerning. It's comforting that both of those are the least high compared to the rest, however, there isn't really any drastic change each year, which is a little concerning in the sense that nothing is being done about it. How can UMD police department act differently or implement new things to break this consistency? 

Also, I noticed for the theft category that in the beginning of the year (January) there was around 305 cases, then mid-year (August) there is a small decline, which then proceeds to rise again by the coming winter. Whilst weather could be a factor, I wonder what caused that change throughout one year. It makes it a little hard to ask questions, since it's not a drastic decrease or increase, but there is a drop so why?  

### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime %>%
  group_by(category) %>%
  mutate(
    total_incidents = sum(Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec), average_incidents = total_incidents / 12
  )
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The categories that have the highest average number of incidents per month are accident, auto stolen, and theft from auto. The lowest averages include homicide, b&e other, and sex offense. I think there is a drastic difference in comparing highest to lowest. Since the highest average is 48% and the lowest being almost 6%. Therefore, the pattern of crime in PG county is high in regards to automobile incidents and car accidents. Personally, as a resident in PG county, this seems accurate since it is a heavy traffic area -- including e-scooters/bikes and pedestrians. 

Also, I'm a little surprised by the average of b&e residential crimes. Again, as a resident living in the College Park area I feel like there are various accounts of b&e throughout dorms or surrounding apartments year round, so I expected higher than 37%. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages %>%
  mutate(
    june_change = ((Jun - average_incidents) / average_incidents) * 100,
    july_change = ((Jul - average_incidents) / average_incidents) * 100,
    august_change = ((Aug - average_incidents) / average_incidents) * 100
  )
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

I've noticed for each crime throughout June it is a little less than the months of July and August -- with July being the month with the highest average. The highest crime in the month of July was robbery vehicle which is a little shocking considering the other summer months are significantly lower, and same goes for cars stolen and then recovered. Recovered vehicles are significantly higher than June and August combined. I find this questionable and would like to dig deeper into why and what it is about July that attracts more vehicles to be recovered after being stolen. 

In regards to the last question, the theory that crime is more likely to occur in the summer months is relatively true, but I'd say more leaning in July. Is it because people are excessively partriotic or something else? 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Lead: July saw a big rise in vehicle-related crime across Prince George's County, outdoing June and August. 

Nut-graph: The data shows that crime was generally lower in June and August but peaked in July, making it the biggest standout of the summer. Vehicle-related crimes were especially noticeable, with vehicle robberies and recovered stolen vehicles reaching their highest average in July. While this supports the idea that crime is more common in the summer, it also raises many questions about what makes July different and why vehicle recoveries are so much higher than in the other summer months.  

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
