Peyton Donaldson
title: "HW3 WH Salaries"
9/11/26

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

After examining the dataframe, my mind immediately went to salary comparison between men and women, especially relative to the White House. Whilst, there are few women that have a salary higher than $170,000 most on the list itself are male. I would dig a bit deeper into this by examining what is the average salary for male v. female WH employees. Also, a major question I'd ask is are men more likely than woman to hold higher paying positions at the White House and if there's a gender pay gap among them. 

Also, the highest salary ranges up to $197,200 per year for senior advisers, deputies, etc. As a journalist, I'd just like some background information on if senior employees are the only ones with the highest salaries and that's why, or if there's more following it. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY)
)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY),
  min_salary = min(SALARY),
  max_salary = max(SALARY)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary is $112,702, while the median is $109,243, which shows that WH salaries are mainly around $109,000-$113,000. The minimum salary is $0 and the maximum is $197,200, showing that there is a pretty large range in-between.The $0 salary is likely the outlier, along with some of the higher salaries, since they are farther from the typical salary represented in the dataset. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I think my best story idea would be to look at whether there is a gender pay gap among White House employees. The next step I would take is to compare the average and median salaries of men and women and then look at their job positions to see if certain roles have bigger salary differences. This could help show whether there is an actual difference in pay or if the difference is mainly because men and women hold different positions OR less women in office, altogether. 
