Peyton Donaldson
title: "HW6_Spotify_Songs"
9/26/26

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify |>
  summarize(mean_danceability = mean(danceability), sd_danceability = sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify |> ggplot(aes(x = danceability)) + geom_histogram(bins=10)
```

### Task 5: Add a line to the histogram for the mean danceability score

Write code to add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify |> ggplot(aes(x = danceability)) + geom_histogram(bins = 10) + geom_vline(xintercept = mean(spotify$danceability))
```

### Task 6: Describe our results

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and distribution tell us about the songs in the dataset?

The danceability scores are mostly concentrated toward the higher end, with many songs having scores around a little over 0.6 to 0.8. The distribution appears to be somewhat skewed to the left because there are fewer songs with very low danceability scores. Overall, the mean and histogram show that songs in this dataset tended to be fairly danceable, although there is still a noticeable range in the scores. 

*I also used AI for Task 4 & 5, because I was confused about how to form a histogram. I know we've done it in class, however, I don't recall exactly how to do it again so I asked for directions on how to. A little more guidance in class would really help, if possible!  

### Task 7: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

